% Paper size, font size, equations align to the left
\documentclass[fleqn,a4paper,11pt]{article}
\title{Anagrams assignment}
\author{Izaak van Dongen}

% Load all the configuration from a package, to keep content separate from
% markup.
\usepackage{mysty}

% The actual document content. Nowadays I would probably put this in a separate
% .tex file and \input that.
\begin{document}
    \maketitle\thispagestyle{empty} % no page number under title
    \tableofcontents

    \section{Introduction and definitions}

    The notion of an anagram is actually quite simple to set-theoretically
    represent. We say two strings \(S\) and \(T\), of lengths \(k\) and \(l\)
    respectively, are palindromes iff
    \begin{equation}
        k = l \land \{x: x \in S\} = \{x: x \in T\}
    \end{equation}
    This capitalises on the fact that sets are unordered. As this seems a
    little terse let's also make the definition
    \begin{equation}
        A = B \iff A \subseteq B \land B \subseteq A \quad
            \text{where} \quad
            C \subseteq D \iff \forall x \in C: x \in D
    \end{equation}
    Similarly to palindromes, some algorithms work better on just letters, and
    this is closer to the notion of what an anagram really is. To avoid having
    "cleanup" code everywhere, this is resolved by saying that the code need
    only behave correctly when supplied with just letters, and when an input
    contains a non-letter, behaviour may be considered undefined and will not
    be tested. An implementation may choose to discard extra letters, or keep
    them.

    \section{Suggested approach by sorting}

    The suggested algorithm was to apply a selection sort to each string. I
    first implemented it like this:

\lstinputlisting[language=Pascal, caption=Initial selection sort]
{../src/Sort.pas}

    This also implements a small optimisation - rather than just searching for
    the minimum each pass, it finds a selects both the maximum and minimum. This
    won't change the complexity of the algorithm, but probably improves the
    constant factor a bit.

    This implementation seemed a little vanilla, so, inspired somewhat by the
    following Haskell:

\lstinputlisting[language=Haskell, caption=Selection sort in Haskell]
{../src/aux/sort.hs}

    I wrote a recursive implementation of the more conventional variation of
    selection sort in Pascal:

\lstinputlisting[language=Pascal, caption=Recursive selection sort in Pascal]
{../src/RecSort.pas}

    However none of this was particularly to much avail, as selection sort has a
    complexity of \(\BigO(n^2)\), owed to its linear number of passes it must
    make. A better idea really would be to implement something like quicksort,
    which would take \(\BigO(n \log(n))\). However, this would in fact also be
    slower than another sorting-based approach, and I'm somewhat constrained for
    time, so I've opted not to do that. As text consists of discrete elements,
    we can apply an even faster class of sorting algorithm - the integer sort.
    These don't rely on comparisons, but instead use integer arithmetic, which
    is generally a lot faster. In this case, the most appropriate would be the
    counting sort, which has a linear runtime in the length of the list.
    Counting sort would be appropriate as it is histogram-based, and we have a
    good restriction on possible characters (ie letters). However, I actually
    won't implement counting sort either because having constructed a histogram,
    we can just \textit{compare histograms}.

    \section{Comparing letter frequencies}

    This approach, for all its speed, is actually pretty simple to implement. A
    multiset, or histogram, of letters can be easily represented as an array of
    integers indexed by letters.

\lstinputlisting[language=Pascal, caption=Basic letter frequencies in Pascal]
{../src/Freqs.pas}

    As the number of letters is constant, the complexity of this program only
    depends on the length of the text, so has complexity \(\BigO(n)\).

    \section{The fundamental theorem of arithmetic}

    Interestingly, as a bit of fun, there is another way to represent a
    histogram. This is as an integer. We say that some histogram represents the
    series of frequencies \(U\), from indices \(1\) to \(k\).  This histogram
    can be encoded as
    \(\prod_{i=1}^{k} P(i)^{U_i}\),
    where \(P\) is the prime function (this is a kind of G\"odel coding). The
    fundamental theorem of arithmetic states that each integer corresponds to a
    unique prime factorisation. This means that each of this prime histogram
    products corresponds to a unique integer. We can then simply perform an
    integer comparison to test equality.  Another benefit of this approach is
    that we can `add' a letter to a histogram simply by multiplying it by the
    corresponding prime number. Here is the implementation:

\lstinputlisting[language=Pascal, caption=Prime-number anagram checking in Pascal]
{../src/Primes.pas}

    It's also impressively short, especially considering that this is written in
    Pascal. A small modification made with regard to the original statement is
    that the correspondence of letters to primes isn't quite linear in the
    progression of primes. I have in fact mapped the most frequently occurring
    letters to the smallest primes. This doesn't have a theoretical effect on
    the algorithm, but it means that in theory the integers being used should
    remain a little smaller.

    This approach does have a slight drawback. For programming languages with
    primarily finite integer types, it may cause integer overflow to occur (this
    is in fact highly likely for longer words, as the value of the integer is
    roughly exponential in length of text). This can lead to false positives.
    Interestingly, it cannot lead to a false negative - this is because really
    an overflowing integer system is a system of modular arithmetic, and
    multiplication is commutative in modular arithmetic, so a weaker version of
    the fundamental theorem still holds..

    \section{Crossing off letters}

    For completeness, I thought I should implement the suggested `crossing off'
    approach. I decided to try and implement it in some semblance of optimality,
    so didn't perform any deletions (I suspect these would be very slow, as they
    require a section of memory to be `shifted'. Instead, I also created a
    boolean array to signify the `crosses'. This is perhaps a nice example of
    space vs time complexity, as the linear auxiliary space here offers a good
    increase in time performance. Despite this, it will still have approximately
    \(\BigO(n^2)\) complexity due to the linear number of linear passes.

\lstinputlisting[language=Pascal, caption=`Crossing off' approach]
{../src/Slow.pas}

    \section{Brute force}

    In my journey from the suggested approach to the linear, histogram based
    approach, I've encountered quite a couple of complexities. I thought that to
    make selection sort feel better, I might implement something even slower.
    That is, brute-forcing permutations.

\lstinputlisting[language=Python, caption=Brute force]
{../src/aux/perms.py}

    We can now add \(\BigO(n!)\) to the collection. Unfortunately, exponential
    complexity remains elusive.

    \section{Testing}

    I, again, wrote a Python script to systematically test my programs.

\lstinputlisting[language=Python, caption=Testing script]
{../src/aux/test_anagrams.py}

    This hardcodes the prescribed test cases, and also provides function to
    compute random anagrams, and random non-anagrams. Anagrams are relatively
    easy - generate a random list and apply a Fisher-Yeates shuffle.
    Non-anagrams can be generated by doing something similar to with palindromes
    - generate an anagram and then apply a mutation to one letter, which
    guarantees that the result will not be an anagram. The program tests these
    categories separately.

    Another fun feature is the use of the carriage return (line 83), to erase
    the previous line of output, meaning it provides a constantly changing
    status. Also of interest, at line 57, the script actually compiles the file
    manually.

    Here we test the first couple of programs:

\begin{lstlisting}[caption=Actual testing]
$ python aux/test_anagrams.py Freqs.pas RecSort.pas Slow.pas Sort.pas
testing Sort.pas     , length 128-256, passes  7035, fails     0
pos:

********************************************************************************
Freqs.pas     passed 100.00% in 0.1383 0.1399 0.1419 0.1438 0.1410 0.1443 0.1472
********************************************************************************

********************************************************************************
RecSort.pas   passed 100.00% in 0.1056 0.1082 0.1111 0.1135 0.1358 0.3291 1.2897
********************************************************************************

********************************************************************************
Slow.pas      passed 100.00% in 0.1403 0.1410 0.1410 0.1456 0.1461 0.1748 0.2377
********************************************************************************

********************************************************************************
Sort.pas      passed 100.00% in 0.1384 0.1422 0.1460 0.1450 0.1491 0.1684 0.2100
********************************************************************************
neg:

********************************************************************************
Freqs.pas     passed 100.00% in 0.1416 0.1426 0.1433 0.1439 0.1445 0.1421 0.1484
********************************************************************************

********************************************************************************
RecSort.pas   passed 100.00% in 0.1063 0.1085 0.1120 0.1138 0.1361 0.3252 1.2903
********************************************************************************

********************************************************************************
Slow.pas      passed 100.00% in 0.1393 0.1416 0.1417 0.1424 0.1480 0.1736 0.2273
********************************************************************************

********************************************************************************
Sort.pas      passed 100.00% in 0.1417 0.1428 0.1441 0.1441 0.1478 0.1686 0.2135
********************************************************************************
\end{lstlisting}
\iffalse $ \fi % syntax highlighting

    Here I've tested all the programs other than `Primes'. When I test primes,
    I get the following:

\begin{lstlisting}[caption=Testing `Primes']
$ python aux/test_anagrams.py Primes.pas
testing Primes.pas   , length 128-256, passes  6048, fails     0
failed on input 'NKYCEOCHALDDNLKPECZREYNQCLZYUFNPZZQBLAAXJDENZOVCQVPXBZMKESUKZQXUVLHTDCIJVAEQFHHWVVCDPJWSIOKIKUVJIGSKPQMDMUDDWFMZYPATESQNTBUYHRZDHBWADIXBCEKSQQYZVOEIXPZLOCIOVLQHDEKEPRAHQCOAEEMYUECGGCPACBRSQXT', 'QEECHKPEIWKDANJLXRQCZNUBZCTMOUCREIETLMCZEZROPOGWIZLQZQUFHNJTPYHDOGAYPXBBNSETCULSVCSKJBAYQAFYLQXECVQKDTVXANQODDHJMSQPDAHOXEZEZCSEGVXBMYLHUPIEISZNVKQCVCKKOHVRIWYIAKDBYAUVLMDEKVWPACDDHQZFPZQDUEC' (b'TRUE\n')
\end{lstlisting}
\iffalse $ \fi % syntax highlighting

    It first successfully passes all the `positive' tests, then passed the
    negative tests until the tests started to happen in the larger length range
    of 128-256, just as I predicted.

    \section{Source}

    All involved files, including this \LaTeX{} document, can be found at
    \url{https://github.com/elterminad0r/anagrams}.


\end{document}
